What makes listeners infer teaching intentions?

### 2022-02-01 paper/manuscript.tex
update methods and results (exp2)
Signed-off-by: Atsuko Tominaga <Tominaga_Atsuko@phd.ceu.edu>

We selected stimuli from our previous experiments (Tominaga et al., \emph{submitted}; Experiment 1). Stimuli were produced by actual pianists on a weighed Yamaha MIDI (Musical Instrument Digital Interface) digital piano and recorded as MIDI files. Multiple pianists played one piece of music with a musical expressive techniques of either articulation (\emph{Figure \ref{fig:stim1}}, A) or dynamics (\emph{Figure \ref{fig:stim1}}, B). Articulation refers to the smoothness of sound, which is comprised of legato and staccato. Legato indicates smooth and connected sound whereas staccato indicates sharp and separate sound. Dynamics refers to the loudness of sound, which is comprised of forte and piano. Forte indicates loud sound while piano indicates soft sound. The piece was taken from ``A Dozen a Day - Play with Ease in Many Keys'' by Edna-Mae Burnam and modified for the experiment. The stimuli were performed around 80 quarter-beats per minute.

@@

For the current experiment, 96 performances were chosen from the valid performances in Tominaga et al.~(\emph{submitted}). We randomly sampled 24 articulation performances and 24 dynamics performances from the teaching condition as well as 24 articulation performances and 24 dynamics performances from the performing condition. It is important to note that the performances from the teaching condition did not necessarily exhibit specific features of teaching that we found in our previous experiments (e.g., exaggeration) since we randomly sampled the performances.

@@

In order to further explore which feature of performance contributed the most to participants' judgments as teaching, multiple regression analyses were conducted. Statistical model assumptions were tested using the \emph{performance} R package (Lüdecke, Ben-Shachar, Patil, Waggoner, \& Makowski, 2021) and all assumptions were met. Since articulation and dynamics consisted of two opposite subcomponents (i.e., legato vs.~staccato, forte vs.~piano) and therefore cannot be summed up to represent each technique, we reported four regression models for each subcomponent (see details in \emph{\protect\hyperlink{dataanalysis}{Data analysis}}).

@@

The experimental setting and procedure were identical to Experiment 1 except that each block consisted of 4 practice trials and 36 experimental trials. The number of trials was reduced due to the time constraint of the experiment.

@@

As Experiment 1, stimuli were produced by actual pianists on a weighed Yamaha MIDI digital piano and recorded as MIDI files. One piece of music was selected as a base piece, which was performed with a musical expressive technique of either articulation (\emph{Figure \ref{fig:stim2}}, A) or dynamics (\emph{Figure \ref{fig:stim2}}, B). The excerpt was taken from ``Sonatina Op.36 (No.3) in C major'' by Muzio Clementi and modified for the experiment. The stimuli were performed around 100 - 120 quarter-beats per minute (i.e., around 250 - 300 ms for 8th-note IOIs).

Recorded piano performances for the stimuli were obtained from our previous experiments (Tominaga et al., \emph{in prep}). Half of the stimuli (36 recordings) was randomly chosen from 248 valid performances in the teaching condition whereas the other half (36 recordings) was randomly chosen from 256 valid performances in the performing condition. In total, 72 recordings were selected as stimuli for the current experiment. Again, it is important to note that each recording from the teaching condition did not necessarily exhibit specific features of teaching performances that we found in the previous experiments since we randomly sampled recordings.

@@

As Experiment 1, stimuli were quantified with regard to tempo (IOIs), articulation (KOT), dynamics (KV) and dynamics contrast (KV-Diff). Only the 8th notes with expressive notations were included for data analysis. We did not run any statistical analysis for Experiment 2.

@@

\hypertarget{scatter-plots}{%
\subsection{Scatter plots}\label{scatter-plots}}

@@

As Experiment 1, it seems that performance tempi (IOIs) are correlated with participants' judgments as teaching. Slower performances were more likely to be judged as teaching. However, the slope of articulation seems to be less steep than that of Experiment 1 (\emph{Figure \ref{fig:ioi-2}}).

@@

Unlike Experiment 1, there seems to be no relationship between KOT values and participants' judgments as teaching. The scatter plot suggests that participants did not use articulation values to infer teaching intentions (\emph{Figure \ref{fig:kot-2}}).

@@

As Experiment 1, there seems to be a relationship between KV values and participants' judgments as teaching (\emph{Figure \ref{fig:vel-2}}). Performances with louder forte adn softer piano seem to be judged as teaching.

@@

As Experiment 1, there seems to be a relationship between KV values and participants' judgments as teaching (\emph{Figure \ref{fig:vel-diff-2}}). Performances with larger contrasts between forte and piano seem to be judged as teaching.

@@

Experiment 2 aims to replicate our previous findings in Experiment 1 with a more naturalistic piece of music. The preliminary data suggest that some features of piano performance contributes to participant' judgments as teaching. For instance, performances with slower tempo were likely to be judged as teaching regardless which expressive technique was implemented. Also, performances with louder forte and softer piano were likely to be judged as teaching for dynamics recordings. Moreover, there seems to be a strong relationship between dynamics contrast between forte and piano and participants' judgments as teaching.

However, it seems there was no relationship between articulation values and participants' judgments as teaching. Compared with Experiment 1, the piece used in Experiment 2 was musically much complex. Therefore, articulation features may not be that reliable in order to infer teaching intentions. Unlike tempo and loudness, which are relatively easier to understand, articulation is very particular to music. The reason we could not see a relationship between articulation values and participants' judgments as teaching might be coming from conceptual and perceptual difficulties of articulation, even for musicians.

>>>>

We selected stimuli from our previous experiments (Tominaga et al., \emph{submitted}). Stimuli were produced by actual pianists on a weighed Yamaha MIDI (Musical Instrument Digital Interface) digital piano and recorded as MIDI files. Multiple pianists played one piece of music with a musical expressive techniques of either articulation (\emph{Figure \ref{fig:stim1}}, A) or dynamics (\emph{Figure \ref{fig:stim1}}, B). Articulation refers to the smoothness of sound, which is comprised of legato and staccato. Legato indicates smooth and connected sound whereas staccato indicates sharp and separate sound. Dynamics refers to the loudness of sound, which is comprised of forte and piano. Forte indicates loud sound while piano indicates soft sound. The piece was taken from ``A Dozen a Day - Play with Ease in Many Keys'' by Edna-Mae Burnam and modified for the experiment. The stimuli were performed around 80 quarter-beats per minute.

@@

For the current experiment, 96 performances were chosen from the valid performances in Tominaga et al.~(\emph{submitted}; Experiment 1). We randomly sampled 24 articulation performances and 24 dynamics performances from the teaching condition as well as 24 articulation performances and 24 dynamics performances from the performing condition. It is important to note that the performances from the teaching condition did not necessarily exhibit specific features of teaching that we found in our previous experiments (e.g., exaggeration) since we randomly sampled the performances.

@@

In order to further explore which feature of performance contributed the most to participants' judgments as teaching, multiple regression analyses were conducted. Statistical model assumptions were tested using the \emph{performance} R package (Lüdecke, Ben-Shachar, Patil, Waggoner, \& Makowski, 2021) and all assumptions were met. Since articulation and dynamics consisted of two opposite subcomponents (i.e., legato vs.~staccato, forte vs.~piano) and therefore cannot be summed up to represent each technique as one value, we reported four separateregression models for each subcomponent (see details in \emph{Data analysis}).

@@

The apparatus and procedure were identical to Experiment 1 except that each block consisted of 4 practice trials and 36 experimental trials. The number of trials was reduced due to the time constraint of the experiment.

@@

As Experiment 1, we selected stimuli from our previous experiments (Tominaga et al., \emph{submitted}). The excerpt was taken from ``Sonatina Op.36 (No.3) in C major'' by Muzio Clementi and modified for the experiment. The excerpt was performed with either articulation (\emph{Figure \ref{fig:stim2}}, A) or dynamics (\emph{Figure \ref{fig:stim2}}, B). The stimuli were performed around 100 - 120 quarter-beats per minute (i.e., around 250 - 300 ms for 8th-note IOIs).

For the current experiment, 96 performances were chosen from the valid performances in Tominaga et al.~(\emph{submitted}; Experiment 2). There were 248 valid performances in the teaching condition and 256 valid performances in the performing condition. We randomly sampled 18 articulation performances and 18 dynamics performances from the teaching condition as well as 18 articulation performances and 18 dynamics performances from the performing condition. In total, 72 recordings were selected as stimuli for the current experiment. Again, it is important to note that each performance from the teaching condition did not necessarily exhibit specific features of teaching that we found in the previous experiments (e.g., exaggeration) since we randomly sampled the performances.

@@

The data analysis was identical to Experiment 1. Only the 8th notes with expressive notations were included for data analysis. As a result, only one 8th note in the 4th measure without any expression was not included.

@@

\hypertarget{correlation-1}{%
\subsection{Correlation}\label{correlation-1}}

As Experiment 1, a Pearson product-moment correlation coefficient was computed by default for correlation analysis and a Spearman's rank correlation coefficient was additionally computed if normality assumption was violated based on the Shapiro-Wilk normality test.

@@

Unlike Experiment 1, there was no significant relationship between performance tempi (IOIs) and participants' judgments as teaching for both techniques (Articulation; \emph{r} = .25, \emph{p} = 0.147, Dynamics; \emph{r} = .39, \emph{p} = 0.019, \emph{Figure \ref{fig:ioi-2}}).

The Shapiro-Wilk normality test revealed that the distribution of IOIs for dynamics recordings was not normally distributed (\emph{p} = 0.080). A Spearman's rank correlation coefficient also showed that there was no significant relationship between tempi for dynamics performances and participants' judgments as teaching (\emph{r} = .29, \emph{p} = 0.091).

@@

Unlike Experiment 1, for articulation recordings, there was no significant relationship between KOT values and participants' judgments as teaching (Legato; \emph{r} = -.15, \emph{p} = 0.390, Staccato; \emph{r} = -.03, \emph{p} = 0.880, \emph{Figure \ref{fig:kot-2}}, left).

For dynamics recordings, there was no significant relationship between KOT values for forte and participants' judgments as teaching (\emph{r} = -.11, \emph{p} = 0.52). However, there was a significant relationship between KOT values for piano and participants' judgments as teaching (\emph{r} = -.35, \emph{p} = 0.03), suggesting that performances with louder staccato were more likely to be considered as teaching performance (\emph{Figure \ref{fig:kot-2}}, right).

@@

As Experiment 1, for dynamics recordings, there was a significant relationship between KV values and participants' judgments as teaching (\emph{Figure \ref{fig:vel-2}}, right). Specifically, performances with louder forte (\emph{r} = .45, \emph{p} = 0.007) and softer piano (\emph{r} = -.45, \emph{p} = 0.01) were more likely to be judged as teaching.

The Shapiro-Wilk normality test revealed that the distribution of KV values for dynamics recordings was not normally distributed (Forte; \emph{p} = 0.048, Piano' \emph{p} = 0.004). Spearman's rank correlation coefficient also confirmed that performances with louder forte (\emph{r} = .43, \emph{p} = 0.010) and softer piano (\emph{r} = -.45, \emph{p} = 0.01) were more likely to be judged as teaching.

For articulation recordings, there was no significant relationship between KV values and participants' judgments as teaching (Legato; \emph{r} = .08, \emph{p} = 0.63, Staccato; \emph{r} = .21, \emph{p} = 0.22, \emph{Figure \ref{fig:vel-2}}, left).

@@

As Experiment 1, for dynamics recordings, there was a significant relationship between KV difference between forte and piano and participants' judgments as teaching (\emph{Figure \ref{fig:vel-diff-2}}, right). Specifically, performances with larger contrasts between forte and piano were more likely to be judged as teaching (From Forte to Piano; \emph{r} = -.75, \emph{p} \textless{} 0.001, From Piano to Forte; \emph{r} = .59, \emph{p} \textless{} 0.001).

The Shapiro-Wilk normality test revealed that the distribution of KV-Diff values for dynamics recordings was not normally distributed (Forte to Piano; \emph{p} = 0.850, Piano to Forte; \emph{p} = 0.008). Spearman's rank correlation coefficient also confirmed that performances with larger contrasts between forte and piano were more likely to be judged as teaching (Forte to Piano; \emph{r} = -.73, \emph{p} \textless{} 0.001, Piano to Forte; \emph{r} = .59, \emph{p} \textless{} 0.001).

\hypertarget{multiple-regression-1}{%
\subsection{Multiple regression}\label{multiple-regression-1}}

As Experiment 1, we performed multiple regression analyses to further explore which feature of performance contributed the most to participants' judgments as teaching. Again, since articulation and dynamics consisted of two opposite subcomponents (i.e., legato vs.~staccato, forte vs.~piano) and therefore cannot be summed up to represent each technique as one value, we reported four separate regression models for each subcomponent (see details in \emph{Data analysis} in Experiment 1).

\hypertarget{legato-1}{%
\subsubsection{Legato}\label{legato-1}}

A multiple regression analysis was conducted to predict participants' judgments as teaching based on performance features of tempo (IOIs), articulation (KOT for legato parts), dynamics (KV for legato parts) and dynamics contrast (KV-Diff from legato to staccato). The result of the regression indicated that the model explained 19.77 \% of the variance (\emph{F}(4, 31) = 3.20, \emph{p} = 0.03). It was found that tempo (IOIs; \emph{\(\beta\)} = 0.58, \emph{p} = 0.011), dynamics (KV) and dynamics contrast (KV-Diff) for the legato parts were articulation for the legato parts were significant predictors of participants' judgments as teaching. However, articulation (KOT; \emph{\(\beta\)} = -0.08, \emph{p} = 0.466) for the legato parts was not a significant predictor as opposed to Experiment 1.

\hypertarget{staccato-1}{%
\subsubsection{Staccato}\label{staccato-1}}

A multiple regression analysis was conducted to predict participants' judgments as teaching based on performance features of tempo (IOIs), articulation (KOT for staccato parts), dynamics (KV for staccato parts) and dynamics contrast (KV-Diff from staccato to legato). The overall was not statistically significant (\(R^{2}\) = 12.28, \emph{F}(4, 31) = 2.20, \emph{p} = 0.09).

\hypertarget{forte-1}{%
\subsubsection{Forte}\label{forte-1}}

A multiple regression analysis was conducted to predict participants' judgments as teaching based on performance features of tempo (IOIs), articulation (KOT for forte parts), dynamics (KV for forte parts) and dynamics contrast (KV-Diff from forte to piano). The result of the regression indicated that the model explained 60.69 \% of the variance (\emph{F}(4, 31) = 14.50, \emph{p} \textless{} 0.001). It was found that dynamics contrast from forte to piano (KV-Diff; \emph{\(\beta\)} = -1.64, \emph{p} \textless{} 0.001) were significant predictors of participants' judgments as teaching.

\hypertarget{piano-1}{%
\subsubsection{Piano}\label{piano-1}}

A multiple regression analysis was conducted to predict participants' judgments as teaching based on performance features of tempo (IOIs), articulation (KOT for piano parts), dynamics (KV for piano parts) and dynamics contrast (KV-Diff from piano to forte). The result of the regression indicated that the model explained 49.49 \% of the variance (\emph{F}(4, 31) = 9.60, \emph{p} \textless{} 0.001). It was found that tempo (IOIs; \emph{\(\beta\)} = 0.55, \emph{p} = 0.022) and dynamics contrast from piano to forte (KV-Diff; \emph{\(\beta\)} = 1.09, \emph{p} \textless{} 0.001) were significant predictors of participants' judgments as teaching.

@@

Experiment 2 aimed to replicate our previous findings in Experiment 1 with a more naturalistic piece of music. We successfully replicated the findings from Experiment 1 in dynamics features. Performances with louder forte and softer piano were likely to be judged as teaching for dynamics recordings. Moreover, there seems to be a strong relationship between dynamics contrast between forte and piano and participants' judgments as teaching.

However, we failed to replicate the results in terms of tempo (IOIs) and articulation (KOT). Compared with Experiment 1, the piece used in Experiment 2 was musically much complex. Therefore, articulation features may not be that reliable in order to infer teaching intentions. Unlike tempo and loudness, which are relatively easier to understand, articulation is very particular to music. The reason we could not see a relationship between articulation values and participants' judgments as teaching might be coming from conceptual and perceptual difficulties of articulation, even for musicians.
